---
title: "Lab 04"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: pdf_document
my_subtitle: "Env Analysis in R"
fancy: true
geometry: margin=1in
latex_engine: pdflatex
colorlinks: true
---

# Lab 04: Raster data management and analysis

### Read the instructions COMPLETELY before starting the lab

This lab builds directly on the raster analysis concepts we worked through in class, including the Week 12 in-class notebook and the Wildcard Thursday activity. You should have that code available for reference.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

## Data

All data for this lab are in the course repository at:

```
../data/wildcard_thursday/erie_raster_ex/
```

You will find two subdirectories:

- `images/` — seven weekly composite GeoTIFF rasters of cyanobacteria bloom intensity in Lake Erie (summer–fall 2017)
- `bounds/` — several shapefiles representing geographic areas of interest within the Lake Erie region

**Before writing any code, read the metadata file at**

**`../data/wildcard_thursday/erie_raster_ex/erie_README.txt`.** 

This is not optional. Understanding your data source, its scaling, and its data key is essential for correct analysis.

## Introduction

Cyanobacteria (blue-green algae) blooms in Lake Erie have become a significant environmental and public health concern. These blooms can produce toxins that contaminate drinking water and harm aquatic ecosystems. The data you will use were produced by NOAA using imagery from the European Space Agency's Sentinel-3 satellite. Each raster represents the maximum observed bloom intensity during a given week.

The product you will be working with is the **CIcyano** (Cyanobacteria Index, cyano only) product. Values in these rasters are stored as **digital numbers (DN)** and must be transformed to meaningful physical units before interpretation. You saw this in class — recall the reverse-scaling formula from the README and from the Week 12 notebook.

A critical aspect of working with remote sensing data is understanding **sentinel values**: special numeric codes that indicate conditions like cloud cover, land pixels, or missing data. These are NOT real observations and must be handled carefully before any analysis.

We begin by loading the relevant packages:

```{r packages, echo=TRUE, message=TRUE}
library(terra)
library(sf)
library(tidyverse)
library(tmap)
library(fs)
```

---

## Task 1: Data management and preprocessing

### 1.1 Inventory your rasters

Use `fs::dir_ls` with the parameter `regexp = "\\.tif$"` to get the file paths for all raster images in the `images/` directory. Store the result in an object called `raster_files`. Print the result — how many files are there?

```{r list_files, echo=TRUE, message=TRUE}
raster_files <- fs::dir_ls(path = "./data/wildcard_thursday/erie_raster_ex/images", regexp = "\\.tif$")
  
raster_files

```

### 1.2 Inspect a single raster

Load the first raster in your list using `terra::rast()`. Examine the following properties and record what you find:

- Coordinate reference system: `terra::crs()`
- Spatial extent: `terra::ext()`
- Spatial resolution: `terra::res()`
- Number of layers: `terra::nlyr()`
- Range of cell values (use `terra::global()` with `"min"` and `"max"`)

```{r inspect_raster, echo=TRUE, message=TRUE}
rast1 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0802_0808.L4.LE3.CIcyano.MAXIMUM_7day.tif")
terra::crs(rast1)
terra::ext(rast1)
terra::res(rast1)
terra::nlyr(rast1)
terra::global(rast1, fun = "min", na.rm = TRUE)
terra::global(rast1, fun = "max", na.rm = TRUE)

```

### 1.3 Write a preprocessing function: `clean_cicyano()`

Looking at the data key in the README, you will notice that values of 0 and 250 and above have special meanings — they are **not** valid bloom observations. Before doing any analysis, these sentinel values must be replaced with `NA`.

Write a function called `clean_cicyano(r)` that:

1. Takes a SpatRaster `r` as input
2. Builds a reclassification matrix that maps **0** and **all values >= 250** to `NA`, and leaves values 1–249 unchanged
3. Applies the reclassification using `terra::classify()`
4. Returns the cleaned raster

*Hint: Your reclassification matrix should handle the no-detection case (0) and all sentinel codes (250 through 255). Review how `terra::classify()` uses a three-column matrix to specify `[from, to, becomes]` intervals. You may find `right = NA` useful here so boundary values like 0 and 250 are included in the intervals.*

```{r clean_function, echo=TRUE, message=TRUE}
clean_cicyano <- function(r) {
  {rclr = matrix(c(0, 0, NA, 1, 249, 1, 250, 252, NA), ncol = 3, byrow = TRUE)} 
  {classified <- terra::classify(r, rcl = rclr, right=NA)}
  classified * r %>% print()} 

rast1.2 <- clean_cicyano(rast1)
#testing
terra::global(rast1.2, fun = "min", na.rm = TRUE)
terra::global(rast1.2, fun = "max", na.rm = TRUE)

```

Test your function on the first raster. Verify it worked by comparing the global min/max of the raw raster vs. the cleaned raster.

### 1.4 Write a transform function: `dn_to_ci()`

The DN values in these rasters are scaled according to the formula in the README. To convert back to physical CI units, use the reverse-scaling formula:

$$CI = 10^{\left(\frac{3.0}{250.0} \times DN - 4.2\right)}$$

Write a function called `dn_to_ci(r)` that takes a cleaned SpatRaster and returns a new raster with values transformed to CI units. You can apply a function to a raster using standard arithmetic operations or by passing a function to the raster.

```{r transform_function, echo=TRUE, message=TRUE}
dn_to_ci <- function(r){
    10**(((3.0 / 250.0) * r) - 4.2)
}

rast1.3 <- dn_to_ci(rast1.2)
terra::global(rast1.3, fun = "min", na.rm = TRUE)
terra::global(rast1.3, fun = "max", na.rm = TRUE)

```

Test your function. What is the range of CI values in the transformed raster? Do they make physical sense given the data key?

### Questions for Task 1:

**Q1.1** What does the CRS of these rasters tell you about how distances and areas are measured? Would you need to reproject before calculating bloom area in square kilometers?
The CRS is in WGS 84 in a specific UTM zone, which means they are already projected for the location. Area is already in meters, so no reprojection is necessary, to obtain sq kilometers, simple math would do.

**Q1.2** Why is it important to remove sentinel values *before* applying the DN-to-CI transform — rather than after? What would happen to a land pixel (value = 252) if you transformed it without cleaning first?
This value would probably be transformed as if it was a meaningful value, and at that point, the range is so complex it would be much harder to exclude from the analyses and mess up results and plots. The real problem is that 252 is not a value representing the algal blooms, so assigning it a CI would be invalid. It is important to remove the sentinel values before anything else because we don't want these values to be analyzed, because there inclusion would be incorrect. Also, it is much easier and efficient to do so at the beginning.

**Q1.3** Why is the `right = NA` argument useful in `terra::classify()` for this task?
Before I used the right=NA, the value of 250 was still being included, even though I was assigning it to be NA. I realized this is because it was interpreting that as values greater than 250 change to NA instead of values greater that AND EQUAL TO 250. Using the 'right=NA' is what told the code to read that as values greater than and equal to 250 like I was intending.

---

## Task 2: Spatial subsetting — crop and mask

### 2.1 Load and align an area of interest

Load the `bb_erie_sandusky` shapefile from the `bounds/` directory using `sf::read_sf()`. Check its CRS. If it does not match the raster CRS, reproject it using `sf::st_transform()`.

```{r load_aoi, echo=TRUE, message=TRUE}

bb <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_sandusky.shp")

sf::st_crs(bb) == sf::st_crs(rast1.3)

```

### 2.2 Crop and mask

Using the **August 16–22, 2017** raster (`ts_2017.0816_0822`):

1. Apply `clean_cicyano()` and `dn_to_ci()` to get a clean, transformed raster
2. Use `terra::crop()` to clip the raster extent to the Sandusky Bay boundary
3. Use `terra::mask()` to set pixels *outside* the boundary to `NA`

Store your final result as `sandusky_0816`.

```{r crop_mask, echo=TRUE, message=TRUE}
rast2 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0816_0822.L4.LE3.CIcyano.MAXIMUM_7day.tif")
rast2.2 <- rast1 %>% clean_cicyano() %>% dn_to_ci()
sandusky_0816 <- terra::crop(rast2.2, bb) %>% terra::mask(., bb, inverse=FALSE)

```

### 2.3 Compare full vs. masked

Using `terra::global()`, calculate and compare the **mean** CI value for:

- The full Lake Erie cleaned/transformed raster
- The Sandusky Bay-masked raster (`sandusky_0816`)

```{r compare_stats, echo=TRUE, message=TRUE}

terra::global(rast2.2, fun = "mean", na.rm = TRUE)
terra::global(sandusky_0816, fun = "mean", na.rm = TRUE)

```

### 2.4 Map it

Create a `tmap` map of `sandusky_0816`. Your map should include:

- A meaningful title
- A legend with appropriate labels
- A scale bar

```{r map_sandusky, echo=TRUE, message=TRUE}

tm_shape(sandusky_0816) + tm_raster(title = "Cyanobacteria Index Values") + tm_title("Cyonabacteria Blooms in the Sandusky Bay") + tm_scalebar(position = c("right","bottom"))
  

```

### Questions for Task 2:

**Q2.1** In your own words, what is the difference between `terra::crop()` and `terra::mask()`? Why did we use both here instead of just one?
Crop focused the map within a particular rectangle surrounding Sandusky. Mask removed the values that weren't within the Sandusky polygon, but still were within the rectangle we cropped to. That is why we used both. We minimized the area we were looking at, but also removed values that were within the area but not inside the portion we actually wanted to focus on.

**Q2.2** The mean CI value for Sandusky Bay is likely different from the full lake mean. What could explain this? Does this necessarily mean Sandusky Bay has more or less cyanobacteria than average?
The average CI was greater when looking at just Sandusky. This would mean that the SanduskyBay area has more cyanobacteria than the average. We see different average values here because one area (Sandusky) does not represent the entire dataset. So outside of the Sanduaky Bay area, there are likely more small values (indicating low cyanobacteria) relative to the amount of greater values, than what we see in Sandusky.

---

## Task 3: Multi-temporal comparison

### 3.1 Load and stack multiple rasters

Load and clean **at least four** of the seven rasters (choose dates that span a meaningful time range). Apply `clean_cicyano()` and `dn_to_ci()` to each. Then combine them into a single multi-layer SpatRaster using `c()`.

*Tip: using `purrr::map()` over your `raster_files` vector, combined with your preprocessing functions, can make this more efficient.*

```{r stack_rasters, echo=TRUE, message=TRUE}

ras0802 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0802_0808.L4.LE3.CIcyano.MAXIMUM_7day.tif")%>% clean_cicyano() %>% dn_to_ci()
ras0823 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day.tif")%>% clean_cicyano() %>% dn_to_ci()
ras0920 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0927_1003.L4.LE3.CIcyano.MAXIMUM_7day.tif")%>% clean_cicyano() %>% dn_to_ci()
ras1025 <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.1025_1031.L4.LE3.CIcyano.MAXIMUM_7day.tif")%>% clean_cicyano() %>% dn_to_ci()

layeredR <- c(ras0802, ras0823, ras0920, ras1025)

```

### 3.2 Temporal summary

Use `terra::global()` on your multi-layer raster to calculate the **mean** CI value per layer (use `na.rm = TRUE`). This will give you one mean value per date.

Organize the results into a data frame with at minimum two columns: `date` and `mean_ci`. You will need to extract the date from each filename. Use `basename()` and/or `stringr::str_extract()` to parse the date string from the file path — the date range is the part of the filename that looks like `0802_0808`.

```{r temporal_summary, echo=TRUE, message=TRUE}
means <- terra::global(layeredR, fun = "mean", na.rm = TRUE)
dates <- stringr::str_sub(basename(raster_files), "9", "17") %>% .[-c(2,5,6)]

df <- data.frame(date = dates, mean_ci = means)
df

```

### 3.3 Plot the time series

Using `ggplot2`, create a line or point plot showing mean CI values over time. Label your axes clearly. Which date in your dataset shows the highest mean bloom intensity? 
0823_0829 has the highest.

```{r time_series_plot, echo=TRUE, message=TRUE}

ggplot(df) + geom_point(aes(date, mean), cex = 3) + labs(x = "Dates", y = "Mean Cyanobacteria Index") + theme_minimal()


```

### 3.4 Binary bloom comparison

Choose any **two dates** from your dataset. For each, create a binary raster where cells with valid CI values are coded as `1` and all other cells (NA) as `0`. You will need to think about how to do this using `terra::classify()` or logical raster operations.

Use these binary rasters to create two new rasters:

- **Persistent bloom:** cells where blooms were present on *both* dates
- **New bloom:** cells where blooms were present on the *later* date but NOT the earlier date

Map both results.

*This requires careful logical thinking before writing code. Plan your approach in pseudocode first.*

```{r binary_comparison, echo=TRUE, message=TRUE}

earlier <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0802_0808.L4.LE3.CIcyano.MAXIMUM_7day.tif")
later <- terra::rast("./data/wildcard_thursday/erie_raster_ex/images/ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day.tif")

rcl = matrix(c(0, 0, 0, 1, 249, 1, 250, 256, 0), ncol = 3, byrow = TRUE)
earlierbin <- terra::classify(earlier, rcl = rcl, right=NA)
laterbin <- terra::classify(later, rcl = rcl, right=NA)

persistent_bloom <- earlierbin * laterbin
new_bloom <- laterbin - earlierbin

plot(persistent_bloom)
plot(new_bloom)

rcl2 = matrix(c(-1, 0, NA, 1, 1, 1), ncol = 3, byrow = TRUE)
new_bloom2 <- terra::classify(new_bloom, rcl = rcl2, right=NA)

```

### Questions for Task 3:

**Q3.1** What are the limitations of using **mean CI value** as a measure of bloom severity across the whole lake? What would be a more robust alternative metric?
Using the mean values across the whole lake does not tell us the bloom severity of specific areas. We could utilize maximum if we want to focus on the worst of the blooms.

**Q3.2** Stacking rasters with `c()` in terra creates a multi-layer SpatRaster. What requirement must all rasters meet in order to be stacked this way? (Think: extent, resolution, CRS.)
They must all have the same extent, resolution, and CRS. They would have to have the same origin and cell shape and position in order to line up correctly.

---

## Task 4: Zonal analysis with areas of interest

### 4.1 — Load all boundary files

Use `list.files()` to get all `.shp` files from the `bounds/` directory. Load each with `sf::read_sf()`, then combine them into a single sf object using `dplyr::bind_rows()`. Make sure each polygon has an identifier — if not, add a column with the filename or AOI name.

```{r load_all_aois, echo=TRUE, message=TRUE}
list.files(path = "./data/wildcard_thursday/erie_raster_ex/bounds", pattern = "\\.shp$")

a <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_islands.shp")
b <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_maumee.shp")
c <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_sandusky_plus.shp")
d <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_large.shp")
e <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_sandusky.shp")
f <- sf::read_sf("./data/wildcard_thursday/erie_raster_ex/bounds/bb_erie_west_sister.shp")

shps <- dplyr::bind_rows(a,b,c,d,e,f)
  
shps$name <- c("islands", "maumee", "sandusky_plus", "large", "sandusky", "west_sister")

shps

```

### 4.2 Extract by area

Using the date you identified in Task 3.3 as having the highest mean bloom intensity, extract CI values for each area of interest using `terra::extract()`. Calculate the **mean** and **maximum** CI per AOI.

*Tip: `terra::extract()` returns a data frame. The `ID` column links rows to the AOI that was used for extraction. You can use `group_by()` and `summarize()` to aggregate.*

```{r zonal_extract, echo=TRUE, message=TRUE}

a1 <- terra::extract(ras0823, a) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
b1 <- terra::extract(ras0823, b) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
c1 <- terra::extract(ras0823, c) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
d1 <- terra::extract(ras0823, d) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
e1 <- terra::extract(ras0823, e) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
f1 <- terra::extract(ras0823, f) %>% group_by(ID) %>% summarize(CImean = mean(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE), CImax = max(ts_2017.0823_0829.L4.LE3.CIcyano.MAXIMUM_7day, na.rm=TRUE))
a1
b1
c1
d1
e1
f1

```

### 4.3 Summary table

Present your results as a clean summary table. Which AOI had the highest mean CI value on that date? Which had the highest maximum?
Maumee had the highest mean and maximum CI.

```{r summary_table, echo=TRUE, message=TRUE}

summary <- dplyr::bind_rows(a1,b1,c1,d1,e1,f1)
  
summary$name <- c("islands", "maumee", "sandusky_plus", "large", "sandusky", "west_sister")

summary

```

### 4.4 Map with AOI overlays

Create a `tmap` map showing the CI raster for your chosen date with the AOI boundaries overlaid. Use `tm_borders()` for the boundaries and label them if possible.

```{r map_with_aois, echo=TRUE, message=TRUE}
tm_shape(ras0823) + tm_raster() +
  tm_shape(a) + tm_borders(lwd=1.5)  +
  tm_shape(b) + tm_borders(lwd=1.5)  +
  tm_shape(c) + tm_borders(lwd=1.5)  +
  tm_shape(d) + tm_borders(lwd=1.5)  +
  tm_shape(e) + tm_borders(lwd=1.5)  +
  tm_shape(f) + tm_borders(lwd=1.5) 

```

### Questions for Task 4:

**Q4.1** You may have gotten `NA` values in your `terra::extract()` results for some AOIs. What could cause this? Name at least two reasons.
One reason could be that areas within the polygon did not exist in the raster data. Another reason could be that some of the area in that location was land and therefore does not have a CI value.

**Q4.2** Looking at the spatial distribution of your AOIs on the map, which areas do you expect to be most prone to severe blooms, and why? (Consider geography, hydrology, land use in the Lake Erie watershed.)
The area of Lake Erie that is within the AOIs in the part near Toledo, Ohio. The areas most prone to severe blooms are the shore lines of the lake. This is because the shores are closer to where people live, and human activities such as fertilizing their lawns result in excess nutrients being broughout into the lake through runoff, causing eutrophication and algal/cyanobacteria blooms.

---


## Questions

1. Throughout this lab you have been working with **raster** data. In what ways was the analysis workflow different from working with vector data (like you did in Labs 1–3)? What concepts translated directly, and what was fundamentally different?
I think the most different thing was that working with a raster wasn't like working with a vector dataframe. In a way it was more simple, as I didn't have to use column names and that sort of thing. But many functions worked very similar, they just had a different name, like tm_raster, for example.

2. Cyanobacteria blooms are affected by temperature, nutrient runoff, and lake circulation. Based on the temporal patterns you observed in Task 3, does the timing of peak bloom match what you would expect ecologically? Explain your reasoning (I know you're not an expert on blooms - just demonstrate thinking).
My highest mean CI was in late August, and my lowest one was in late October. I would expect that more sever blooms are correlated with higher temperatures, as august has higher average temperatures than in october. august may also have more rainfall, which would increase the amount of runoff bringing excess nutrients into the lake. 

3. We used `terra::extract()` in Task 4 to perform a zonal summary. How does this relate conceptually to the **zonal operations** category of map algebra discussed in class?
For zonal operations, we apply functions to multiple raster cells, but we are dealing with multiple zones that were defined by a category, the AOI in this case. 

